### 2024-09-15 actuarial.qmd
20240915 - optimism and overconfidence

In general, many so-called experts are "pseudo-experts" who do not know the boundaries of their competence—that is, they do not know what they do not know; they have the illusion of validity of their predictions and are [overconfident](#sec-cognitiveBiasesOverconfidence) about their predictions [@Kahneman2011].
Yet, many people arrogantly proclaim to have predictive powers, including in low-validity environments such as fantasy football.
Indeed, pundits are more likely to be television guests if they are opinionated, clear, and (overly) confident and make big, bold predictions, because they are more entertaining [even though they tend to be less accurate than individuals whose thinking is more complex and less decisive; @Kahneman2011; @Silver2012].
True experts know their limits in terms of knowledge and ability to predict.

>>>>

In general, many so-called experts are "pseudo-experts" who do not know the boundaries of their competence—that is, they do not know what they do not know; they have the illusion of validity of their predictions and are [overconfident](#sec-cognitiveBiasesOverconfidence) about their predictions [@Kahneman2011].
Yet, many people arrogantly proclaim to have predictive powers, including in low-validity environments such as fantasy football.
Indeed, pundits are more likely to be television guests if they are opinionated, clear, and (overly) confident and make big, bold predictions, because they are more entertaining [even though they tend to be less accurate than individuals whose thinking is more complex and less decisive; @Kahneman2011; @Silver2012].
[Optimism](#sec-cognitiveBiasesOptimism) and [(over)confidence](#sec-cognitiveBiasesOverconfidence) are valued by society [@Kahneman2011].
Nevertheless, true experts know their limits in terms of knowledge and ability to predict.

### 2024-09-15 cognitive-bias.qmd
20240914 - base rates
20240915 - optimism and overconfidence

*Overplacement* involves believing that one is better than others or will perform better than others, even when they do not.
For instance, it is a common finding that more than half of people believe they are "above average" (i.e., above the median), even though that is statistically impossible.
This calls to mind the fictitious Lake Wobegon in the radio show *A Praririe Home Companion*, "where all the women are strong, all the men are good-looking, and all the children are above average."

@@

The confidence that people tend to have in their predictions depends less and the quantity and quality of the evidence for their predictions and more on the quality of the narrative they can tell about the (little) information they have immediate access to [@Kahneman2011].
For instance, people tend not to consider other relevant, important factors that they do not have immediate access to, consistent with the [WYSIATI heuristic](#sec-heuristicsWYSIATI).
Instead, people tend to focus on the coherence and plausibility of the story, which differs from its probability [@Kahneman2011].
That is, one's confidence is not necessarily related to the likelihood that the judgment is correct—it often reflects the "illusion of validity" [@Kahneman2011].
As one gains more knowledge in a domain, the person also tends to become unrealistically overconfident in their ability to make accurate judgments and predictions [@Kahneman2011].

@@

Players' performance in fantasy football, and human behavior more generally, is complex and multiply determined (i.e., is influenced by many factors).
Despite the bluster of so-called experts who pretend to know more than they can know, no one can consistently and accurately predict how all players will perform.
Remain humble in your predictions; do not be more confident than is warranted.
If you approach the task of prediction with humility, you may be more able to be flexible and more willing to consider other players who you can draft for good value.

@@

*Optimism bias* is the tendency for people to overestimate one's likelihood of experiencing positive events and underestimate one's likelihood of experiencing negative events.
For instance a manager might overestimate the likelihood that their team will win the championship and may underestimate the likelihood that they may lose a player to injury.

@@

The *base rate fallacy* (aka base rate neglect) is the tendency to ignore information about the general probability of an event in favor of specific information about the event.
The *base rate* is a marginal probability, which is the general probability of an event irrespective of other things.
Focusing on the specific information about the event (and thus underweighting the base rate) involves the [representativeness heuristic](#sec-heuristicsRepresentativeness) [@Kahneman2011]—making judgments based on how similar the event is to a mental prototype regardless of how frequent the event the event actually is.
In general, people tend to overestimate the likelihood of low base-rate events [@Kahneman2011].

>>>>

*Overplacement* involves believing that one is better than others or will perform better than others, even when they do not.
For instance, it is a common finding that more than half of people believe they are "above average" (i.e., above the median), even though that is statistically impossible.
This calls to mind the fictitious Lake Wobegon in the radio show *A Praririe Home Companion*, "where all the women are strong, all the men are good-looking, and all the children are above average."
As an example, 90% of drivers believe they are better than average [@Kahneman2011].

@@

The confidence that people tend to have in their predictions depends less and the quantity and quality of the evidence for their predictions and more on the quality of the narrative they can tell about the (little) information they have immediate access to [@Kahneman2011].
For instance, people tend not to consider other relevant, important factors that they do not have immediate access to, consistent with the [WYSIATI heuristic](#sec-heuristicsWYSIATI).
Instead, people tend to focus on the coherence and plausibility of the story, which differs from its probability [@Kahneman2011].
That is, one's confidence is not necessarily related to the likelihood that the judgment is correct—it often reflects the "illusion of validity" [@Kahneman2011].
As one gains more knowledge in a domain, the person also tends to become unrealistically overconfident in their ability to make accurate judgments and predictions [@Kahneman2011].
In general, people tend to focus on what they know and to neglect what they do not know leading to overconfidence in one's beliefs [@Kahneman2011].

@@

People tend to focus on the role of skill and to neglect the role of luck when explaining the past and predicting the future, giving people an illusion of control [@Kahneman2011]. 
Players' performance in fantasy football, and human behavior more generally, is complex and multiply determined (i.e., is influenced by many factors).
Despite the bluster of so-called experts who pretend to know more than they can know, no one can consistently and accurately predict how all players will perform.
Remain humble in your predictions; do not be more confident than is warranted.
If you approach the task of prediction with humility, you may be more able to be flexible and more willing to consider other players who you can draft for good value.

@@

*Optimism bias* is the tendency for people to overestimate one's likelihood of experiencing positive events and underestimate one's likelihood of experiencing negative events.
For instance a manager might overestimate the likelihood that their team will win the championship and may underestimate the likelihood that they may lose a player to injury.
In terms of its key consequences for decisions, optimism bias may be considered the most important cognitive bias [@Kahneman2011].
In general, people show an optimism bias.
However, some people ("optimists") are even more optimistic than others [@Kahneman2011].
The optimism bias—in particular a delusional optimism—can lead to [overconfidence](#sec-cognitiveBiasesOverconfidence) [@Kahneman2011].
Although optimism is associated with many advantages in terms of well-being, it can also lead to excessive risk-taking and to costly persistence [@Kahneman2011].

@@

The *base rate fallacy* (aka base rate neglect) is the tendency to ignore information about the general probability of an event in favor of specific information about the event.
The *base rate* is a marginal probability, which is the general probability of an event irrespective of other things.
Focusing on the specific information about the event (and thus underweighting the base rate) involves the [representativeness heuristic](#sec-heuristicsRepresentativeness) [@Kahneman2011]—making judgments based on how similar the event is to a mental prototype regardless of how frequent the event the event actually is.
As noted by @Kahneman, "people who have information about an individual case rarely feel the need to know the statistics of the class which the case belongs." (p. 249).
In general, people tend to overestimate the likelihood of low base-rate events [@Kahneman2011].
